---
title: "simulate_true_rkf_biomass"
output: html_document
date: "2025-09-26"
---

load libraries
```{r}
library(dplyr)
library(ggplot2)

set.seed(1234) # for reproducibility 
```

create spatial grid - later on could substitute this for GOA grid 
```{r}
# grid <- expand.grid(x=1:20, y=1:20)
# 
# #randomly assign 30% of grid as untrawlable - could substitute this for GOA T/UT maps 
# grid <- grid %>%
#   mutate(bottom_type = rbinom(nrow(grid), size = 1, prob = 0.3)) # 0 = trawlable; 1 = untrawlable 
# 
# #simulate biomass of species x 
# #set base biomass in trawlable habitat - could update with realistic biomass 
# t_biomass <- 5 
# #set habitat multiplier - again update with realistic ratios for each species 
# ut_multiplier <- 4
# 
# grid <- grid %>%
#   mutate(biomass = t_biomass * ifelse(bottom_type == 1, ut_multiplier, 1) * rlnorm(nrow(grid), meanlog = 0, sdlog = 0.3))
# ## probably need to put more thought into the random noise (i.e. the rlnorm() part...)
```

plot simulated biomass grid 
```{r}
# ggplot(grid, aes(x = x, y = y, fill = biomass)) +
#   geom_tile() +
#   scale_fill_viridis_c() +
#   #facet_wrap(~bottom_type, labeller = labeller(bottom_type = c("0" = "Trawlable", "1" = "Untrawlable"))) +
#   theme_minimal()
```

other simulated biomass grid based on levels of untrawlable to trawlable and adds prob of presence 
```{r}
grid <- expand.grid(
  x = 1:20,
  y = 1:20
)

#add continuous trawlable/untrawlable variable
n <- nrow(grid)

# Number of cells in each group - have 70% be trawlable (1 to 5) and 30% be untrawlable (-4 to 0)
n_trawlable <- round(n * 0.7)
n_untrawlable <- n - n_trawlable

# Sample values for each group
trawlable_vals <- sample(1:5, n_trawlable, replace = TRUE)
untrawlable_vals <- sample(-4:0, n_untrawlable, replace = TRUE)

# Combine and shuffle
bottom_type <- sample(c(trawlable_vals, untrawlable_vals))

# Assign to grid
grid <- grid %>%
  mutate(bottom_type = bottom_type)

# 3. Define probability of presence based on habitat
# We'll scale probabilities so habitat = -4 → 4× more likely than habitat = 5
# Use logistic function for probability: p = 1 / (1 + exp(beta * habitat))
# Solve beta such that p(-4)/p(5) ≈ 4

# Let's approximate beta
# p(-4)/p(5) = (1 / (1 + exp(beta*-4))) / (1 / (1 + exp(beta*5))) ≈ 4
# Simplify: (1 + exp(5*beta)) / (1 + exp(-4*beta)) ≈ 4
# We'll solve numerically

beta_fun <- function(beta) (1 + exp(5*beta)) / (1 + exp(-4*beta)) - 4
beta <- uniroot(beta_fun, c(-5, 5))$root

# 4. Calculate probability of presence
grid <- grid %>%
  mutate(
    prob_presence = 1 / (1 + exp(beta * bottom_type)))

# 5. Simulate presence/absence
grid <- grid %>%
  mutate(
    presence = rbinom(nrow(grid), size = 1, prob = prob_presence))

# 6. Simulate biomass where present
base_biomass <- 5
ut_multiplier <- 4

grid <- grid %>%
  mutate(
    biomass = ifelse(presence == 1, base_biomass * ut_multiplier * rlnorm(nrow(grid), meanlog = 0, sdlog = 0.3), 0))
```

```{r}
ggplot(grid, aes(x = x, y = y, fill = bottom_type)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal()

ggplot(grid, aes(x = x, y = y, fill = biomass)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal()
```

```{r}
# total true biomass of species 
sum(grid$biomass)
```

simulate sampling 10% of the area, but only in the trawlable zones (1-5) with a net 
and simulate sampling eDNA from those same cell plus 10% more that are in untrawlable 
```{r}
trawlable_cells <- grid %>%
  filter(bottom_type > 0)
untrawlable_cells <- grid %>%
  filter(bottom_type <= 0)

trawlable_sampled_cells <- trawlable_cells %>% 
  sample_frac(0.1)
untrawlable_sampled_cells <- untrawlable_cells %>% 
  sample_frac(0.1)

# 3. Simulate observed biomass with sampling variability
# Assume net catches a proportion of the true biomass with some noise
# For example: observed ~ rnorm(mean = true_biomass, sd = 0.2*true_biomass)
trawlable_sampled_cells <- trawlable_sampled_cells %>%
  mutate(
    observed_biomass = pmax(0, rnorm(n(), mean = biomass, sd = 0.2 * biomass)) # no negative biomass
    )

all_sampled_cells <- trawlable_sampled_cells %>%
  bind_rows(untrawlable_sampled_cells)

# Let's assume:
# eDNA_conc = alpha * biomass + random_noise
# alpha = conversion factor from biomass to DNA units (e.g., copies/mL)
alpha <- 50  

# Add variability in eDNA detection (e.g., sampling & PCR)
# Use lognormal noise to ensure positive values
all_sampled_cells <- all_sampled_cells %>%
  mutate(
    eDNA_conc = rlnorm(n(), meanlog = log(alpha * biomass + 1e-3), sdlog = 0.3)
  )
```

what's the DNA concentration ratio of untrawlable to trawlable 
```{r}
# Define habitat classes
# Trawlable: habitat > 0
# Untrawlable: habitat <= 0
all_sampled_cells <- all_sampled_cells %>%
  mutate(class = ifelse(bottom_type > 0, "trawlable", "untrawlable"))

# Calculate mean eDNA in each habitat type
eDNA_summary <- all_sampled_cells %>%
  group_by(class) %>%
  summarise(mean_eDNA = mean(eDNA_conc, na.rm = TRUE))

# Calculate the ratio
eDNA_ratio <- eDNA_summary$mean_eDNA[eDNA_summary$class == "untrawlable"] / eDNA_summary$mean_eDNA[eDNA_summary$class == "trawlable"]

# View results
eDNA_summary
eDNA_ratio
```

if the trawlable cells were thought to be representative of the entire survey grid... 
```{r}
# estimated biomass based on trawling 
sum(all_sampled_cells$observed_biomass, na.rm = T) / 0.1
```

if it was known that trawlable cells make up 70% of the area and that untrawlable made up remaining but at a higher biomass (according to eDNA ratio)... 
```{r}
# estimated biomass 
((sum(all_sampled_cells$observed_biomass, na.rm = T) / 0.1)*0.7) + (sum(all_sampled_cells$observed_biomass, na.rm = T) / 0.1)*eDNA_ratio*0.3
```

based on this, catchability (q) would be: 
```{r}
density_in_trawlable <- sum(all_sampled_cells$observed_biomass, na.rm = T) / 0.1
density_in_untrawlable <- sum(all_sampled_cells$observed_biomass, na.rm = T) / 0.1*eDNA_ratio
area_trawlable <- 0.7 
area_untrawlable <- 0.3

q <- density_in_trawlable*(area_trawlable+area_untrawlable)/((density_in_trawlable*area_trawlable)+(density_in_untrawlable*area_untrawlable))
q
```



### now ask gemini to write me a bootstrapped simulation code to do this many times 

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# --- ADJUSTABLE PARAMETERS ---
# Set the proportion of habitat that is trawlable
p_trawlable <- 0.7 

# Set the proportion of cells to be sampled
p_sample <- 0.2 
# -----------------------------


# Define a function to run a single simulation with given parameters
run_simulation <- function(prop_trawlable, sample_prop) {
  
  # 1. SET UP THE VIRTUAL WORLD
  grid <- expand.grid(x = 1:20, y = 1:20)
  n <- nrow(grid)
  
  # Create trawlable and untrawlable habitat based on the parameter
  n_trawlable <- round(n * prop_trawlable)
  n_untrawlable <- n - n_trawlable
  trawlable_vals <- sample(1:5, n_trawlable, replace = TRUE)
  untrawlable_vals <- sample(-4:0, n_untrawlable, replace = TRUE)
  grid$bottom_type <- sample(c(trawlable_vals, untrawlable_vals))
  
  # 2. SIMULATE THE FISH POPULATION
  beta <- -0.4625 # Pre-calculated beta
  grid <- grid %>%
    mutate(prob_presence = 1 / (1 + exp(beta * bottom_type)),
           presence = rbinom(n, size = 1, prob = prob_presence))
  
  base_biomass <- 5
  ut_multiplier <- 4
  grid <- grid %>%
    mutate(
      biomass = ifelse(presence == 1, 
                       base_biomass * ifelse(bottom_type <= 0, ut_multiplier, 1) * rlnorm(n, meanlog = 0, sdlog = 0.3), 
                       0)
    )
  
  true_biomass <- sum(grid$biomass)
  
  # 3. SIMULATE THE SURVEYS
  trawlable_cells <- grid %>% filter(bottom_type > 0)
  untrawlable_cells <- grid %>% filter(bottom_type <= 0)
  
  # Sample a fraction of each habitat type based on the parameter
  trawlable_sampled_cells <- trawlable_cells %>% sample_frac(sample_prop)
  untrawlable_sampled_cells <- untrawlable_cells %>% sample_frac(sample_prop)
  
  trawlable_sampled_cells <- trawlable_sampled_cells %>%
    mutate(observed_biomass = pmax(0, rnorm(n(), mean = biomass, sd = 0.2 * biomass)))
  
  all_sampled_cells <- bind_rows(trawlable_sampled_cells, untrawlable_sampled_cells)
  
  alpha <- 50
  all_sampled_cells <- all_sampled_cells %>%
    mutate(eDNA_conc = rlnorm(n(), meanlog = log(alpha * biomass + 1e-3), sdlog = 0.3))
    
  # 4. CALCULATE BIOMASS ESTIMATES
  eDNA_summary <- all_sampled_cells %>%
    mutate(class = ifelse(bottom_type > 0, "trawlable", "untrawlable")) %>%
    group_by(class) %>%
    summarise(mean_eDNA = mean(eDNA_conc, na.rm = TRUE))
  
  ratio_df <- tidyr::spread(eDNA_summary, class, mean_eDNA)
  eDNA_ratio <- ifelse(nrow(ratio_df) > 0 && "trawlable" %in% names(ratio_df) && ratio_df$trawlable > 0, 
                       ratio_df$untrawlable / ratio_df$trawlable, 0)
  
  total_observed_biomass <- sum(trawlable_sampled_cells$observed_biomass, na.rm = TRUE)
  
  # Estimate 1: Trawl-only (extrapolated using sample proportion)
  trawl_only_estimate <- total_observed_biomass / sample_prop
  
  # Estimate 2: Trawl + eDNA (using habitat and sample proportions)
  prop_untrawlable <- 1 - prop_trawlable
  biomass_from_trawlable <- (total_observed_biomass / sample_prop) * prop_trawlable
  biomass_from_untrawlable <- (total_observed_biomass / sample_prop) * eDNA_ratio * prop_untrawlable
  trawl_edna_estimate <- biomass_from_trawlable + biomass_from_untrawlable

  return(c(
    true_biomass = true_biomass,
    trawl_only = trawl_only_estimate,
    trawl_plus_eDNA = trawl_edna_estimate
  ))
}

# Run the simulation 1,000 times with the specified parameters
set.seed(42) # For reproducible results
n_reps <- 1000
results_matrix <- replicate(n_reps, run_simulation(prop_trawlable = p_trawlable, sample_prop = p_sample))

# Process results into a tidy data frame for plotting
results_df <- as.data.frame(t(results_matrix)) %>%
  pivot_longer(
    cols = c(trawl_only, trawl_plus_eDNA),
    names_to = "method",
    values_to = "estimate"
  )

# Calculate the average true biomass across all simulations
mean_true_biomass <- mean(results_df$true_biomass)

# 5. PLOT THE RESULTS
ggplot(results_df, aes(x = estimate, fill = method)) +
  geom_density(alpha = 0.6) +
  geom_vline(aes(xintercept = mean_true_biomass), color = "black", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("trawl_only" = "#1f78b4", "trawl_plus_eDNA" = "#33a02c"),
                    labels = c("Trawl-Only Estimate", "Trawl + eDNA Estimate")) +
  labs(
    title = "Distribution of Biomass Estimates from 1,000 Simulations",
    subtitle = paste0("Trawlable Habitat: ", p_trawlable * 100, "%, Sample Intensity: ", p_sample * 100, "%"),
    x = "Estimated Biomass",
    y = "Density",
    fill = "Estimation Method"
  ) +
  annotate("text", x = mean_true_biomass, y = 0, label = "Mean True\n Biomass", vjust = -0.5, hjust = -0.1, color = "black") +
  theme_minimal()
```

next try a model-based estimation of biomass based on sampled data (i.e. tinyVAST)

use the data frame set up above 
```{r}
ggplot(all_sampled_cells, aes(x = x, y = y, fill = observed_biomass)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal()

ggplot(all_sampled_cells, aes(x = x, y = y, fill = eDNA_conc)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal()
```

fit the model 
```{r, results = 'hide'}
library(tinyVAST)
library(fmesher)
library(sf)

sim_data <- all_sampled_cells
sim_data$species <- "species_A"

# make mesh
mesh = fm_mesh_2d(grid[,c('x','y')], 
                   cutoff = 0.1 )
mesh = fm_refine(mesh)
plot(mesh)

# separate spatial sd
#sem = "
#  species_A <-> species_A, SD # sd_species_A
#  species_B <-> species_B, SD # sd_species_B
#  species_C <-> species_C, SD # sd_species_C
#"

# define distribution 
Family = list(
    "species_A" = tweedie(link="log")
  )

# Formula for simulation with site replicates (i.e. adding in random effect of site)
#formula = observed_biomass ~ 0 + species
formula = observed_biomass ~ 1

# add error_level column
sim_data$error_level <- sim_data$species
 
# fit model
out = tinyVAST( data = sim_data,
                formula = formula,
                spatial_domain = mesh,
                #space_term = sem,
                family = Family,
                space_columns = c("x", "y"),
                #variable_column = "species",
                distribution_column = "error_level",
                control = tinyVASTcontrol(
                  trace = 1,
                  calculate_deviance_explained = FALSE
                ) )
```

look at model output
```{r}
out
```

predict biomass 
```{r}
# Make prediction grid (all spatial locations you want predictions for)
pred_grid <- grid[,c(1:2)] %>%
  mutate(species = "species_A",
         error_level = "species_A")

# Predict expected biomass
pred_out <- predict(
  out,
  newdata = pred_grid,
  #type = "response",  # "response" gives back biomass scale, not link scale
  what = "p_g",
  se.fit = TRUE
)

# Add predictions to grid
pred_grid$predicted_biomass <- pred_out$fit
pred_grid$predicted_biomass_se <- pred_out$se.fit

pred_grid$area <- 1

# Integrate the fitted model to get total biomass 
total_biomass <- integrate_output(out,
                                  newdata = pred_grid,
                                  area = pred_grid$area,
                                  bias.correct=FALSE,
                                  apply.epsilon=FALSE)

total_biomass
```





